Insight Into Lumbar Back Pain \subtitle{What the Lumbar Spine Tells About Your Life}

### 2014-09-03 draft.tex
Most of the variables  no italic

Our contributions are:
\begin{itemize}
\item An IVA-framework for back pain analysis based on image-derived variables of 2,240 subjects \com{Ist diskutabel, so richtig interaktiv ist es nicht. IVA wurde auch noch nicht vorher definiert.}
%%
\item The identification and evaluation of lumbar spine shape properties for back pain diagnosis
%%
\item Detection of correlations between image-based and socio-demographic as well as medical parameters for an improved hypothesis generation
%%
\item Identification of the most important variables via data mining methods (including a novel semi-quantitative evaluation concept) embedded in our IVA-framework \com{Wieder: IVA wurde noch nicht definiert}
%%
\end{itemize}

% -- Old Introduction
% Epidemiology is the study of causation of diseases.
% %%
% Large population studies, such as the Study of Health in Pomerania (SHIP) \ref{SHIP} gather as much information as possible about participants to be assessed towards different diseases.
% %%
% These information are used to determine risk factors for diseases, helping people to make their lifestyle healthier or helping in diagnosing a disease.
% %%
% Epidemiological research is strongly hypothesis driven.
% %
% Observations made by clinicians are translated into hypothesis, which are then statistically evaluated using data variables from epidemiological studies.
%
% Modern cohort studies often comprise medical image data.
% %%
% \com{Short summary}
% %%
%
% Back pain is one of the most frequent diseases in the western civilization.
% %%
%
% Our goal is to combine data mining algorithm with data visualization to provide insight into the quality of image derived data to analyze if it acts as a risk factor for a disease.
% %\com{Combine Data Mining Algo}
%
% More than 5 variables are rarely analyzed simultaneously.
% %%
% %The variables analyzed simultaneously is limited to
% Our contributions are:
% \begin{itemize}
% 	\item Analyzing back pain using image-derived variables of 2,240 subjects.
% 	\item Assessing the suitability of lumbar spine shape for diagnosing back pain
% 	\item Analyzing correlations between between image-based and socio-demographic as well as medical parameters.
% 	\item Identification of most important variables using data mining methods \com{noch schoen schreiben}
% 	\item semiquantitative Auswertmöglichkeiten werden dem User oft zur verfügung gestellt
% 	\item Techniken sind als Teil des IVA-Frameworkes zu verstehen
% \end{itemize}

\section{\uppercase{Epidemiological Background}}
\label{sec:EpidemiologicalBackground}
%\noindent \com{Background. Epidemiological Workflow, focus on statistical resilience, image data hard to analyze due to the large amount of subjects, poor image quality and lack of methods}.
\noindent Epidemiological reasoning relies on a strict statistically driven workflow \cite{Fletcher}:
%%
\begin{itemize}
%%
	\item Physicians formulate hypotheses based on observations made in their clinical practice.
%%
	\item To assess a hypothesis, epidemiologists compile a list of variables depicting it.
%%
	\item Statistical methods, such as regression analysis, assess the association of selected variables with the investigated disease.
%%
\end{itemize}
%%
Mutually dependent variables make this analysis challenging.
%%
Many diseases, such as different cancer types, are more likely with increasing age.
%%
When for example analyzing influences of nutrition to prostate cancer, the results need to be age-normalized.
%%
Age acts as a \emph{confounder} for prostate cancer.
%%
%These \emph{confounding} variables are often hard to find.
%%
Statistical correlation does not imply causation--epidemiologists need to assess the medical soundness of the statistical results.
%%------------------------------
\subsection{Epidemiological Data}
Epidemiological data originates from a wide range of studies.
%%
The study type depends on the condition of interest.
%%
Most common are case-control studies, analyzing one specific disease and its influences on the human body.
%%
We focus on data from large scale cohort studies.
%%
These studies aim to collect as much data as possible for each subject.
%%
As a result, these data can be analyzed regarding many diseases and conditions.
%%

@@

no group (a), subdivision into male and female subjects (b), quartiles of age (c) and quartiles of body size (d).

@@

The analysis of the dummy variable yields similar results like the blood pressure medication.
%%
Variables, such subjects sized 139-164 cm, between 64 and 90 years of age or nutrition related parameters are dominantly populated by one manifestation.
%%
The classifier neglects the other group and yields a error rate below $0.5$.
%The dummy variable analysis yields fewer results.
%%
%It shows that subjects between 64 and 90 years of age can be detected with 23\% error rate by using the curvature angle and coronal mean curvature.
%%
%Subjects with body size between 177 cm and 202 cm can classified with an error rate of 22,1\%. solely using mean curvature.
%%
\subsubsection{Gender Groups}
%%
Classification using groups divided by gender do not produce satisfying results.
%%
Only hypothyroidism could be described for male subjects with an error rate of $0.24$ for 110 subjects using the \emph{mean curvature} and \emph{curvature angle}.
%%
Since there are only 30 male subjects diagnosed with hypothyroidism, the statistical power of the result is reduced.
%%
The dummy variable analysis showed that female subjects of 139-164 cm body size could be discriminated using the mean curvature and curvature angle, with an error of $0.38$.

%\com{Potentially skip these due to the confounding of body size}
\subsubsection{Age Groups}
%%
Gender could be classified for each age group using \emph{mean curvature} and \emph{curvature angle}.
%%
The accuracy varies between $0.35$ ($43.8-54$ years of age, 3 decision rules) to $0.27$ ($21-43.8$ years of age, 6 decision rules).
%%
The starting age of smoking could be characterized well with an error rate between $0.25$ to $0.3$ for all age groups except of $21.1-43.8$ years of age.
%%
The result however is probably over fitted to the data due to tree sizes between $16$ and $23$.
%%
\com{Feedback der Greifswalder notwendig, warum das so sein koennte.}
%%

Many variables, such as body size can be described with an error rate of $0.35$ to $0.45$ but only using large decision trees with over 20 rules.
%Body-size can be determined with an error rate of $0,36$ to $0,39$ for subject
% age groups, but also imposes over fitting to the data yielding a decision tree of over 30 rules.
%%
Notable is also the increasing accuracy with increasing decision tree size.
%%
For subjects between $43.8-54$ years of age, the classifier discriminated subjects with thyroid nodules with an error rate of $0.32$.

@@

\subsubsection{Size Groups}
%%
Many previously described results are confounded with subject size.
%%
Differences between male and female subjects are explained by to the average body size difference.
%%
For example, large subjects are already characterized by their rather straight spine.
%
The question is, whether the inter-group spine-variability parameter is enough for predicting other parameter or not.
%%
Dividing subjects into body size groups potentially highlights classifications not confounded by body size.
%%
\paragraph{Large Decision Trees. }
Back pain-associated variables can be explained for various size-groups, but we could not extract universal rules.
%%
Radiating back pain could be described with error rate of $0.39$ using 23 rules for subjects between $171-177~cm$ body size using \emph{torsion} and \emph{mean curvature}.
%%
For subjects sized $177-202~cm$ the accuracy drops to $0.47$ using 20  decision rules.
%%
There are several decision trees for laboratory values, such as alanine aminotransferase value (relevant for diagnosis of lover or gallbladder illness) in the blood can be described with an error rate of $0.4$ ($139-164~cm$) to $0.36$ ($164 - 171~cm$).
%%
Similar values can be observed for cholesterol or age.
%%
Due to the large decision trees, these results are not usable and impose overfitting to the data.
%%
\paragraph{Small Decision Trees. }
%%
The dummy variables show several variables described using only one decision rule with accuracy between $0.42$ to $0.47$.
%%
Most of these variables have a dominant manifestation and the classifier shows a low detection precision for the second manifestation.
%%
These variables include nutrition parameter, thyroid disorder and social problems induced by back pain.

@@

\section{\uppercase{Conclusion}}
\label{sec:Conclusion}
\noindent The presented results indicate that the parameters torsion, curvature and curvature angle are not enough to characterize pathological changes in the lumbar spine.
%%
They are well suited to characterize subject gender due to their body size difference using mean curvature and mean curvature angle.
%%
This is due to the higher curvature of smaller subjects and the significant difference of average body size between males and females.
%%
Gender was characterized well in the analysis of all subjects as well as subjects divided by age.

Subjects grouped using body size have a similar spine shape and a lower variance.
%%
The remaining information within these groups is not enough to characterize back pain-related variables.
%%
%Certain parameters can be classified in 
For some body size groups, certain nutrition or psychological parameters can be predicted with a comparatively high error rate.
%Certain subgroups, however, can be classified for various size groups, such as nutrition or psychological parameters.
%%

@@

\subsection{Summary and Future Work}
%%
\noindent We showed, that torsion, curvature and curvature angle of the lumbar spine at the presented precision are not enough to characterize lumbar back pain in the SHIP data set.
%%
Our method made it possible to assess their discriminative power, which is largely limited for separating male and female subjects, nutrition variables, as well as different disease indicators.
%%
The C4.5 algorithm showed to be an effective tool for evaluating a set of derived metrics regarding their suitability to classify non-image parameters.
%%
Over-fitting to the data indicated by complex decision trees have to be taken into account into the analysis as well.
%%

>>>>

Our contributions are:
\begin{itemize}
\item An IVA-framework for back pain analysis based on image-derived variables of 2,240 subjects \com{Ist diskutabel, so richtig interaktiv ist es nicht. IVA wurde auch noch nicht vorher definiert.}
%%
\item The identification and evaluation of lumbar spine shape properties for back pain diagnosis
%%
\item Detection of correlations between image-based and socio-demographic as well as medical parameters for an improved hypothesis generation
%%
\item Identification of the most important variables via data mining methods (including a novel semi-quantitative evaluation concept) embedded in our IVA-framework \com{Wieder: IVA wurde noch nicht definiert}
%%
\end{itemize}
%% ---------------------------------------------
\section{\uppercase{Epidemiological Background}}
\label{sec:EpidemiologicalBackground}
%%
\noindent Epidemiological reasoning relies on a strict statistically driven workflow \cite{Fletcher}:
%%
\begin{itemize}
%%
	\item Physicians formulate hypotheses based on observations made in their clinical practice.
%%
	\item To assess a hypothesis, epidemiologists compile a list of variables depicting it.
%%
	\item Statistical methods, such as regression analysis, assess the association of selected variables with the investigated disease.
%%
\end{itemize}
%%
Mutually dependent variables make this analysis challenging.
%%
Many diseases, such as different cancer types, are more likely with increasing age.
%%
When for example analyzing influences of nutrition to prostate cancer, the results need to be age-normalized.
%%
Age acts as a \emph{confounder} for prostate cancer.
%%
%These \emph{confounding} variables are often hard to find.
%%
Statistical correlation does not imply causation--epidemiologists need to assess the medical soundness of the statistical results.
%%------------------------------
\subsection{Epidemiological Data}
Epidemiological data originates from a wide range of studies.
%%
The study type depends on the condition of interest.
%%
Most common are case-control studies, analyzing one specific disease and its influences on the human body.
%%
We focus on data from large scale cohort studies.
%%
These studies aim to collect as much data as possible for each subject.
%%
As a result, these data can be analyzed regarding many diseases and conditions.
%%

@@

no group (a), subdivision into \emph{male and female subjects} (b), quartiles of \emph{age} (c) and quartiles of \emph{body size} (d).

@@

The analysis of the dummy variable yields similar results like the \emph{blood pressure medication}.
%%
Variables, such as subjects sized 139-164 cm, between 64 and 90 years of age or nutrition related parameters are dominantly populated by one manifestation.
%%
The classifier neglects the other group and yields a error rate below $0.5$.
%The dummy variable analysis yields fewer results.
%%
%It shows that subjects between 64 and 90 years of age can be detected with 23\% error rate by using the curvature angle and coronal mean curvature.
%%
%Subjects with body size between 177 cm and 202 cm can classified with an error rate of 22,1\%. solely using mean curvature.
%%
\subsubsection{Gender Groups}
%%
Classification using groups divided by gender do not produce satisfying results.
%%
Only \emph{hypothyroidism} could be described for male subjects with an error rate of $0.24$ for 110 subjects using the \emph{mean curvature} and \emph{curvature angle}.
%%
Since there are only 30 male subjects diagnosed with \emph{hypothyroidism}, the statistical power of the result is reduced.
%%
The dummy variable analysis showed that female subjects of \emph{139-164~cm body size} could be discriminated using the \emph{mean curvature} and \emph{curvature angle}, with an error of $0.38$.

%\com{Potentially skip these due to the confounding of body size}
\subsubsection{Age Groups}
%%
Gender could be classified for each age group using \emph{mean curvature} and \emph{curvature angle}.
%%
The accuracy varies between $0.35$ (\emph{$43.8-54$ years of age}, 3 decision rules) to $0.27$ (\emph{$21-43.8$ years of age}, 6 decision rules).
%%
The starting age of smoking could be characterized well with an error rate between $0.25$ to $0.3$ for all \emph{age} groups except of \emph{$21.1-43.8$ years of age}.
%%
The result however is probably over fitted to the data due to tree sizes between $16$ and $23$.
%%
\com{Feedback der Greifswalder notwendig, warum das so sein koennte.}
%%

Many variables, such as \emph{body size} can be described with an error rate of $0.35$ to $0.45$ but only using large decision trees with over 20 rules.
%Body-size can be determined with an error rate of $0,36$ to $0,39$ for subject
% age groups, but also imposes over fitting to the data yielding a decision tree of over 30 rules.
%%
Notable is also the increasing accuracy with increasing decision tree size.
%%
For subjects between \emph{$43.8-54$ years of age}, the classifier discriminated subjects with thyroid nodules with an error rate of $0.32$.

@@

\subsubsection{Size Groups}
%%
Many previously described results are confounded with subject size.
%%
Differences between \emph{male} and \emph{female} subjects are explained by to the average \emph{body size} difference.
%%
For example, large subjects are already characterized by their rather straight spine.
%
The question is, whether the inter-group spine-variability parameter is enough for predicting other parameter or not.
%%
Dividing subjects into \emph{body size} groups potentially highlights classifications not confounded by \emph{body size}.
%%
\paragraph{Large Decision Trees. }
Back pain-associated variables can be explained for various \emph{size}-groups, but we could not extract universal rules.
%%
Radiating back pain could be described with error rate of $0.39$ using 23 rules for subjects between \emph{$171-177~cm$ body size} using \emph{torsion} and \emph{mean curvature}.
%%
For subjects sized \emph{$177-202~cm$} the accuracy drops to $0.47$ using 20 decision rules.
%%
There are several decision trees for laboratory values, such as \emph{alanine aminotransferase} value (relevant for diagnosis of liver or gallbladder illness) in the blood can be described with an error rate of $0.4$ (\emph{$139-164~cm$}) to $0.36$ (\emph{$164 - 171~cm$}).
%%
Similar values can be observed for \emph{cholesterol} or \emph{age}.
%%
Due to the large decision trees, these results are not usable and impose overfitting to the data.
%%
\paragraph{Small Decision Trees. }
%%
The dummy variables show several variables described using only one decision rule with accuracy between $0.42$ to $0.47$.
%%
Most of these variables have a dominant manifestation and the classifier shows a low detection precision for the second manifestation.
%%
These variables include \emph{nutrition} parameter, \emph{thyroid disorder} and \emph{social problems induced by back pain}.

@@

\section{\uppercase{Conclusion}}
\label{sec:Conclusion}
\noindent The presented results indicate that the parameters \emph{torsion}, \emph{curvature} and \emph{curvature angle} are not enough to characterize pathological changes in the lumbar spine.
%%
They are well suited to characterize subject gender due to their body size difference using \emph{mean curvature} and \emph{mean curvature angle}.
%%
This is due to the higher curvature of smaller subjects and the significant difference of average \emph{body size} between \emph{males} and \emph{females}.
%%
Gender was characterized well in the analysis of all subjects as well as subjects divided by \emph{age}.

Subjects grouped using \underline{body size} have a similar spine shape and a lower variance.
%%
The remaining information within these groups is not enough to characterize back pain-related variables.
%%
%Certain parameters can be classified in 
For some \emph{body size} groups, certain \emph{nutrition} or \emph{psychological} parameters can be predicted with a comparatively high error rate.
%Certain subgroups, however, can be classified for various size groups, such as nutrition or psychological parameters.
%%

@@

\subsection{Summary and Future Work}
%%
\noindent We showed, that \emph{torsion}, \emph{curvature} and \emph{curvature angle} of the lumbar spine at the presented precision are not enough to characterize lumbar back pain in the SHIP data set.
%%
Our method made it possible to assess their discriminative power, which is largely limited for separating \emph{male} and \emph{female} subjects, \emph{nutrition} variables, as well as different disease indicators.
%%
The C4.5 algorithm showed to be an effective tool for evaluating a set of derived metrics regarding their suitability to classify non-image parameters.
%%
Over-fitting to the data indicated by complex decision trees have to be taken into account into the analysis as well.
%%
